---
title: "Homework 12: Population Scale Analysis"
author: "Nicolò (PID: A18109144)"
format: pdf
---

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes

```{r}
# read the file
pop <- read.table("rs8067378_ENSG00000172057.6.txt")
head(pop)
```

```{r}
nrow(pop)
```

Sample size for each genome

```{r}
table(pop$geno)
```

A/A genotype median expression level

```{r}
aa <- pop[which(pop$geno == "A/A"),]
head(aa)
median(aa$exp)
```

A/G genotype median expression level

```{r}
ag <- pop[which(pop$geno == "A/G"),]
head(ag)
median(ag$exp)
```

G/G genotype median expression level

```{r}
gg <- pop[which(pop$geno == "G/G"),]
head(gg)
median(gg$exp)
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)

ggplot(pop) + aes(geno, exp, fill = geno) +
  geom_boxplot(notch = TRUE)
```

By looking at the boxplot, it seems like the SNP does affect the ORMDL3 gene expression level, in particular the G allele seems to lower the expression level.