---
title: "Class 06: R Functions"
author: "Nicolò (PID: A18109144)"
format: pdf
---

# All about functions in R

Functions are the way we get stuff done in R. We call a function to read data, compute stuff, plot stuff, etc. etc.

R makes writing functions accessible but we should always start by trying to get a working snippet of code first before we write our function.

## Today's lab

We will grade a whole class of student assignments. We will always try to start with a simplified version of the problem.

```{r}
# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
```

```{r}
mean(student1)
```

Let's be nice instructors and drop the lowest score so the answer here should be 100

```{r}
min(student1)
```

I found the `which.min()` function that may be useful here. How does it work? Let's just try it:

```{r}
student1
which.min(student1)
```

I can use the minus syntax trick to get everything but the element with the min value

```{r}
student1[-which.min(student1)]
```

I have my first working snippet of code

```{r}
mean(student1[-which.min(student1)])
```

Let's test on the other students

```{r}
mean(student2[-which.min(student2)])
mean(student3[-which.min(student3)])
```

Where is the problem? It's the NA value, let's ignore it

```{r}
mean(student2, na.rm=TRUE)
```

No bueno. We need to fix this!

I want to stop working with `student1`, `student2` etc. and typing it out every time so let's instead work wirh an input called `x`.

```{r}
x <- student2
x
```

We want to overwrite the NA values with zero - if you miss a homework you score zero on this homework.

Google and Claude told me about the is.na() function, let's use it.

```{r}
is.na(x)
```

We can use logicals to index a vector

```{r}
x[is.na(x)] <- 0
x
mean(x[-which.min(x)])
```

This is my working snipped of code that solves the problem for all my example student inputs

```{r}
x <- student2
#Mask NA values to 0
x[is.na(x)] <- 0
#Drop lowest score and get the mean
mean(x[-which.min(x)])
```

> Q1. Write a function grade() to determine an overall grade from a vector of student homework assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an NA value) this can be used as a score to be potentially dropped. Your final function should be adquately explained with code comments and be able to work on an example class gradebook such as this one in CSV format: “https://tinyurl.com/gradeinput” [3pts]

```{r}
grade <- function(x) {
  #Mask NA values to 0
  x[is.na(x)] <- 0
  #Drop lowest score and get the mean
  mean(x[-which.min(x)])
}
```

Use this function:

```{r}
grade(student1)
grade(student2)
grade(student3)
```

We need to read the gradebook

```{r}
gradebook <- read.csv("https://tinyurl.com/gradeinput", row.names = 1)
gradebook
```

Calculate the averages for every student in the gradebook:

```{r}
avg_scores <- apply(gradebook, 1, grade)
avg_scores
```

> Q2. Using your grade() function and the supplied gradebook, Who is the top scoring student overall in the gradebook? [3pts]

```{r}
which.max(apply(gradebook, 1, grade))
```

> Q3. From your analysis of the gradebook, which homework was toughest on students (i.e. obtained  the lowest scores overall? [2pts]

```{r}
mask <- gradebook

mask[is.na(mask)] <- 0
hw.ave <- apply(mask, 2, mean)
hw.ave
which.min(hw.ave)
```

> Q4. Optional Extension: From your analysis of the gradebook, which homework was most predictive of overall score (i.e. highest correlation with average grade score)? [1pt]

```{r}
highest_cor <- apply(mask, 2, cor, y=avg_scores)
highest_cor
which.max(highest_cor)
```